Sex-Specific Evolution of the Meiotic Recombination Rate

### 2020-07-30 doc/MS_v6.tex
overdue commits

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Sex-Specific Evolution of the Meiotic Recombination Rate},
  pdfauthor={April L. Peterson, Bret Payseur},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{titling}

\pretitle{%
  \begin{center}
  \LARGE
  \includegraphics[width=4cm,height=6cm]{logo.png}\\[\bigskipamount]
}
\posttitle{\end{center}}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\title{Sex-Specific Evolution of the Meiotic Recombination Rate}
\author{April L. Peterson, Bret Payseur}
\date{2020-06-14}

@@

Although meiotic recombination is required for successful gametogenesis
in most species that reproduce sexually, the rate of crossing over
varies among individuals. \textbf{define genome-wide recombination
rate}.

Differences in recombination rate between females and males are perhaps
the most striking form of this variation. Existing data fail to directly
address the extent to which recombination experiences similar
evolutionary pressures. To fill this gap, we measured meiotic
recombination in both sexes for a panel of house mouse cross three
subspecies. Using inbred strains and single cell immunohistochemistry
allowed us to place sex-specific observations within the same genetic
background and meiotic context. Our results indicate highly discordant
evolutionary patterns in the two sexes. Whereas male recombination rates
show evidence of rapid evolution over short evolutionary timescales,
female recombination rates measured in the same strains are mostly
static. These results strongly indicate that house mouse has two genome
wide recombination rates which display distinct evolutionary
trajectories.

@@

\hypertarget{importance-of-recombination-rate}{%
\subsubsection{0. Importance of recombination
rate}\label{importance-of-recombination-rate}}

Meiosis can be reduced to the expression of (2n -\textgreater{} 4n
-\textgreater{} 2n -\textgreater{} 1n) which tracks the duplication of a
diploid genome into haploid cell products. The meiotic program relies on
crossovers and the process of recombination to ensure the correct
separation of chromosomes. The total number of COs per cell (at the 4n
stage) is equal to the genome wide recombination rate (gwRR).
\textbf{The genome wide recombination rate is a cell based metric for
genome wide recombination rate is tightly connected to organisms fitness
through correct chromosome segregation and fertility.}

\textbf{at the fine scale / in regards to genetic variation}, In the
indriect manner, the recombination rate regulates populations' response
to selection, and determine the fate of novel mutations by transferring
beneficial mutations onto novel genetic backgrounds or by breaking
linkage of negative mutations from beneficial genetic backgrounds
(cite). This process shapes the genomic patterns of genetic variation,
with high recombination areas of genome having more nucleotide variation
while areas of low recombination have decreased genetic variation (Begun
Aquadro, Nachman, Payseur).

\textbf{but back to the genome-wide recombination rate}, The process of
chromosome segregation is integral to the process of producing haploid
cells and can be considered a constraint in regards to the evolution.
When the genome wide recombination rate is framed within the single cell
context it is closely connected to the contraints which are imposed at
the single cell level. Reflecting the importance of it's role in
chromosome segreation is that the number of haploid pairs is strong
predictor for genome wide RR across large species(taxa) / divergence
scales (Otto and Payseur, 2019; Stapley et al., 2017). (the number of
crossovers per chromosome are highly conserved)\textless However, whie
the rate of recombination is highly variable, the count of crossovers
per chromosome is conserved between 1 to 3 (cite), suggesting that such
thresholds/constraints may be behind the overall patterns.\textgreater{}

\textbf{connection of single cell based gwRR to thresholds imposed at
the single cell level} (why is gwRR the better scale for evolutionary
patterns?)

\hypertarget{understanding-levels-of-variation-in-gwrr}{%
\subsubsection{1. Understanding levels of variation in
gwRR}\label{understanding-levels-of-variation-in-gwrr}}

Because we know the meiotic programs, we have expectations for where the
thresholds(constraints) of the gwRR the rate will fall. First, the lower
threshold is defined as an obligate crossover per homolog pair which
promotes proper chromosome segregation (Nagaoka et al., 2012). second
(while there is less evidence,), the upper threshold maybe
regulated/shaped/affected due to rates of ectopic recombination or the
programmed DNA damage preceding crossovers act as constraints to the
upper threshold of meiotic recombination (cite). Yet, Within these
thresholds the genome wide recombination rate still varies across
species and indiviuals. which is the subject of an open question in the
field. \textbf{the gwRR is the best metric for since both of these
constraints are thought to act at the cell wide level.}

\textbf{Sex is one of the most notable axes along which individuals
vary}. Sexual dimorphism in recombination rates is called heterochiasmy
(Lenormand, 2003). An understanding of how sex shapes the evolution of
recombination cannot be achieved with available data. Comprehensive
comparisons of female and male recombination rates usually come from
outbred populations humans (Kong et al., 2004,, 2008, 2014; Halldorsson
et al., 2019), dog, cattle (Ma et al., 2015; Shen et al., 2018), (other
sheep) sheep (Johnston et al., 2016), and mouse (CC cite) in which the
role of sex is confounded with the contributions of genetic variation.
Under anisogamy where a species has distinct gametes, there is no first
principle which would predict the evolution of sexually dimorphic
recombination rates. Yet heterochiasmy, is commonly observed in
dioecious species, suggesting that other meiotic traits which
distinguish the gametes and their meiotic programs, for example
symmetrical vs asymmetrical cell division, may impose selection for
sexually dimorphic recombination rates. Although it is clear that the
relationship between female and male recombination rates can differ
among species, comparisons between and within closely related species
are missing. Direct contrasts between females and males across a common,
diverse set of genomic backgrounds would reveal whether the
recombination rate evolves differently in the sexes.

\textless It's important to note, -- that the genome wide recombination
rate can be decomposed -- and connected to related traits (which also
show conserved patterns of sexual dimorphism\textgreater{}

\textbf{understanding heterochiasmy and sexual dimorphism in meiosis
have advanced since first being documented (morgan 1914) -- more than
just the direction -- more evidence is accumulating that there are
conserved patterns of sexual dimorphism in meiotic recombinatio -
related to recombination rate}. Recently two specific features have been
highlighted as conserved patterns of heterochiasmy:

Cahoon and Libuda (2019) show In species with meiotic chromatin (SC axis
length) quantified for both sexes, longer chromosome axis generally has
higher recombination rates. Sardell and Kirkpatrick (2020), A survey of
51 species shows there are conserved broad scale recombination landscape
differences. Generally males have telomere-bias crossover placement and
females have more uniform placement.

These observations raise the question of conservation in sexually
patterns are maintained with the gwRR and it's decomposed traits

\textbf{remember these points kinda distract from gwRR aspect}

\hypertarget{the-house-mouse-is-a-great-model-for-uncovering-evolutionary-patterns-at-a-short-timescale}{%
\subsubsection{2. The House Mouse is a great model for uncovering
evolutionary patterns at a short
timescale}\label{the-house-mouse-is-a-great-model-for-uncovering-evolutionary-patterns-at-a-short-timescale}}

The House mouse complex arose from a recent radiation providing an
opportunity to interrogate natural variation at short evolutionary
scales, potentially narrowing the mutational space for identifying
variants which would cause gwRR variation. Add divergence ranges.
divergence spanning 0.5 mya to 5 mya.

Classical lab strains of mice have generated extensive knowledge central
to meiosis and outcomes on recombination through extensive studies on
the genes involved in the meiotic recombination pathway (cite) and
previous crosses for understanding the genetic architecture of
recombination rate variation ({\textbf{???}}); Murdoch et al. (2010);
Wang and Payseur (2017){]}.

\textbf{Wild derived inbred strains generate the best comparison of
females and males, besides the sex chromosomes, the mouse for each
genome are almost identical}. Inbred strains enable one of the most
direct comparison of both female and male -- versions of meiosis
--enable use to conclude that sex is a primary factor --

Additionally due to their global distribution allows us to compare
samples of natural diversity from a broad geographic range. (wild
derived inbred strains -- will capture genetic diversity not represented
in classical lab strains.)

While there are strains with Robertsonian translocations, in house mouse
and related murid species it is possible to assemble sets of strains
with identical karyotypes, 20 pairs of acrocentric chromosomes.

House mouse is well suited for single cell cytology approaches. The
single cell level allows integration of data at a closer connection to
the molecular pathway and meiotic program. In addition to the
quantification of gwRR. This study quantifies precursors to crossovers,
double strand breaks (DSBs), and meiotic chromosome morphology across
stages of meiosis which would not be accessible to with genetic linkage
data and crosses.

\hypertarget{what-we-accomplished-in-this-paper}{%
\subsubsection{3. What we accomplished in this
paper}\label{what-we-accomplished-in-this-paper}}

For the first time in many of these strains, We report a rare, direct,
evolutionary comparison of recombination rate in females and males for
this short evolutionary scale. We use rare strains from a broad range of
geographic locations of the species territory.

The gwRR is made up of crossovers occurring on individual chromosomes
within cells. Thus we quantified meiotic chromosome morphology (SC
length) and placement of crossovers to comprise an approximate picture
of the recombination landscape.

Our results indicate rapid male specific evolution of gwRR occurred in a
subset of house mouse lines. We observed up to a 30\% difference
(translating into approximately \textasciitilde7 more crossovers per
cell), a surprising amount considering the short evolutionary time
scale.

@@

We used a panel of wild-derived inbred strains of house mice (\emph{Mus
musculus}) and related murid species to profile natural genetic
variation in recombination (Table 1). Our survey included 5 strains from
\emph{M. m. musculus}, 4 strains from \emph{M. m. domesticus}, 2 strains
from \emph{M. m. molossinus}, 2 strains from \emph{M. m. castaneus}, and
1 strain each from \emph{M. spicilegus} and \emph{M. spretus}.
\textless removed \emph{M. caroli}\textgreater{} We subsequently denote
strains by their abbreviated subspecies and name
(e.g.~\emph{domesticus\textsuperscript{WSB}}). Mice were housed at
dedicated, temperature-controlled facilities in the UW-Madison School of
Medicine and Public Health, with the exception of mice from Gough
Island, which were housed in a temperature-controlled facility in the
UW-Madison School of Veterinary Medicine. Mice from an inbred strain of
Gough Island mice were sampled after \textbf{XX} generations of
brother-sister mating. All mice were provided with ad libitum food and
water. Procedures followed protocols approved by IACUC.

@@

The same dry-down spread technique was applied to both spermatocytes and
oocytes, following (Peters et al., 1997), with adjustment for volumes.
Spermatocyte spreads were collected and prepared as described in
(Peterson et al., 2019). The majority of mice used for MLH1 counts were
between 5 and 12 weeks of age. Juvenile mice between 12 and 15 days of
age were used for DMC1 counts. Both ovaries were collected from embryos
(16-21 embryonic days) or neonates (0-48 hours after birth). Whole
testes were incubated in 3ml of hypotonic solution for 45 minutes.
Decapsulated ovaries were incubated in 300ul of hypotonic solution for
45 minutes. Fifteen microliters of cell slurry (masticated gonads) were
transferred to 80ul of 2\% PFA solution. Cells were fixed in this
solution and dried in a humid chamber at room temperature overnight. The
following morning, slides were treated with a Photoflow wash (Kodak,
cite). Slides were stored at -20*C if not stained immediately. To
visualize the structure of meiotic chromosomes, we used antibody markers
for the centromere (CREST) and lateral element of the synaptonemal
complex (SC) (SYCP3). Crossovers (COs) were visualized as MLH1 foci.
Double strand breaks (DSBs) were visualized as DMC1 foci. The staining
protocol followed (Anderson et al., 1999) and (Koehler et al., 2002).
Antibody staining and slide blocking were performed in 1X antibody
dilution buffer (ADB) (normal donkey serum (Jackson ImmunoResearch), 1X
PBS, bovine serum albumin (Sigma), and Triton X-100 (Sigma)). Following
a 30-minute blocking wash in ABD, each slide was incubated with 60ul of
a primary antibody master mix for 48 hours at 37*C. The master mix
recipe contained polyclonal anti-rabbit anti-MLH1 (Calbiochem; diluted
1:50) or anti-rabbit anti-DMC1 (mix of DMC1), anti-goat polyclonal
anti-SYCP3, (Abcam; diluted 1:50), and anti-human polyclonal antibody to
CREST (Antibodies, Inc; diluted 1:200) suspended in ADB. Slides were
washed twice in 50ml ADB before the first round of secondary antibody
incubation for 12 hours at 37*C. Alexa Fluor 488 donkey anti-rabbit IgG
(Invitrgoen, location; diluted to 1:100) and Coumarin AMCA donkey
anti-human IgG (Jackson ImmunoResearch; diluted to 1:200) were suspended
in ADB. The last incubation of Alexa Fluor 568 donkey anti-goat
(Invitrogen; diluted 1:100) was incubated at 1:100 for 2 hours at 37* C.
Slides were fixed with Prolong Gold Antifade (Invitrogen) for 24 hours
after a final wash in 1x PBS.

@@

Images were captured using a Zeiss Axioplan 2 microscope with AxioLab
camera and AxioVision software (Zeiss, Cambridge, UK). Preprocessing,
including cropping, noise reduction, and histogram adjustments, was
performed using Photoshop (v13.0). Image file names were anonymized
before manual scoring of MLH1 foci or DMC1 foci using Photoshop.

\hypertarget{analysis}{%
\subsubsection{Analysis}\label{analysis}}

\textless new 1/n\textgreater{}

To estimate the number of crossovers across the genome, we counted MLH1
foci. MLH1 foci were counted in cells with intact and complete
karyotypes (19 acrocentric bivalents and XY for spermatocytes; 20
acrocentric bivalents for oocytes) and distinct MLH1 foci. A quality
score ranging from 1 (best) to 5 (worst) was assigned to each cell based
on visual appearance of staining and spread of bivalents. Cells with a
score of 5 were excluded from the final analysis. Distributions of MLH1
count per cell were visually inspected for normality (Supplemental
Figure 1). MLH1 foci located on the XY in spermatocytes were excluded
from counts. In addition to MLH1 counts, we measured several traits to
further characterize the recombination landscape. To estimate the number
of double-strand breaks, a minority of which lead to crossovers, mean
DMC1 foci per cell was quantified for a single male from each of a
subset of strains (\emph{molossinus\textsuperscript{MSM}},
\emph{musculus\textsuperscript{PWD}},
\emph{domesticus\textsuperscript{WSB}}, and
\emph{domesticus\textsuperscript{G}} ). SC morphology and CREST foci
number were used to stage spermatocytes as early zygotene or late
zygotene.

\textless new 2/n\textgreater{} To measure bivalent SC length, two image
analysis algorithms were used. The first algorithm estimates the total
(summed) SC length across bivalents for individual cells (Wang et al.
(2019)). The second algorithm estimates the SC length of individual
bivalents (Peterson et al. (2019)). Both algorithms apply a
`skeletonizing' transformation to synapsed chromosomes that produces a
single, pixel-wide `trace' of the bivalent shape. Total SC length per
cell was quantified from pachytene cell images (Wang et al. (2019)).

To reduce algorithmic errors in SC isolation, outliers were visually
identified at the mouse level and removed from the data set. Mouse
averages were calculated from cell-wide total SC lengths in 3,204 out of
3,884 cells with MLH1 counts. SC length of individual bivalents was
quantified in pachytene cell images (Peterson et al. (2019)). The DNA
CrossOver algorithm (Peterson et al. (2019)) isolates single,
straightened bivalent shapes, returning SC length, location of MLH1
foci, and location of CREST (centromere) foci. The algorithm
substantially speeds the accurate measurement of bivalents, but it
sometimes interprets overlapping bivalents as single bivalents. In our
data set, average proportions of bivalents per cell isolated by the
algorithm ranged from 0.48 (\emph{molossinus\textsuperscript{MSM}} male)
to 0.72 (\emph{musculus\textsuperscript{KAZ}} female). From the total
set of pachytene cell images, 10,222 bivalent objects were isolated by
the algorithm. Following a manual curation, 9,575 single-bivalent
observations remained. The accuracy of the algorithm is high compared to
hand measures after this curation step (Peterson et al., 2019). The
curated single bivalent data supplements our cell-wide MLH1 count data
with MLH1 foci counts for single bivalents. Proportions of bivalents
with the same number of MLH1 foci were compared across strains using a
chi-square test.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Sex-Specific Evolution of the Meiotic Recombination Rate},
  pdfauthor={April L. Peterson, Bret A. Payseur},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{setspace}
\doublespacing
\usepackage{lineno}
\linenumbers
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\title{Sex-Specific Evolution of the Meiotic Recombination Rate}
\author{April L. Peterson, Bret A. Payseur}
\date{2020-07-29}

@@

Although meiotic recombination is required for successful gametogenesis
in most species that reproduce sexually, the rate of crossing over
varies among individuals. Differences in recombination rate between
females and males are perhaps the most striking form of this variation.
Despite the recognition that the sexes recombine at different rates,
existing data fail to address the extent to which recombination
experiences similar evolutionary pressures in the two sexes. To fill
this gap, we measured the genome-wide recombination rate in both sexes
from a panel of wild-derived inbred strains sampled from across three
subspecies of house mice. We applied single-cell immunohistochemistry to
inbred strains to directly compare females and males from the same
genetic background and meiotic context, thereby isolating the effect of
sex on evolutionary differences in recombination. Our results reveal
highly discordant evolutionary trajectories in the two sexes. Whereas
male recombination rates show rapid evolution over short timescales,
female recombination rates measured in the same strains are mostly
static. Strains with high recombination in males have more double-strand
breaks and stronger crossover interference than strains with low
recombination in males, suggesting that these factors contribute to the
sex-specific evolution we document. These results provide the strongest
evidence yet that sex is a primary driver of recombination rate
evolution.

@@

Meiosis converts diploid germ cells into haploid gametes. During meiosis
I, DNA crossovers aid the separation of homologous chromosomes by
physically linking them and establishing tension between them on the
spindle (Petronczki et al., 2003). The wrong number of recombination
events can disrupt chromosomal segregation, leading to infertility,
miscarriage, and birth defects (Hassold and Hunt, 2001). Recombination
also shapes evolution by shuffling the combinations of genetic variants
offspring inherit. Recombination affects the fates of beneficial and
deleterious mutations (Felsenstein, 1974; Fisher, 1930; Hill and
Robertson, 1966) and interacts with natural selection to leave gradients
in genomic patterns of diversity (Begun and Aquadro, 1992; Charlesworth
et al., 1993; Cutter and Payseur, 2013; Nachman and Payseur, 2012; Smith
and Haigh, 1974).

The role of recombination in facilitating meiotic chromosome assortment
suggests that the total number of crossovers in a cell -- the
genome-wide recombination rate -- is an important cellular
characteristic connected to organismal fitness. The dual pressures of
ensuring at least one crossover per chromosome and minimizing levels of
DNA damage and ectopic exchange are thought to impose lower and upper
thresholds on the genome-wide recombination rate (Inoue and Lupski,
2002; Nagaoka et al., 2012). Yet, within these bounds, individuals from
the same species can vary substantially in crossover number (Gruhn et
al., 2013; Johnston et al., 2016; Kong et al., 2008; Ma et al., 2015).

Sex is perhaps the most notable axis along which recombination rate
varies. Broadly speaking, sexual dimorphism in the genome-wide
recombination rate assumes two forms. In species such as
\emph{Drosophila melanogaster}, one sex completes meiosis without
forming crossovers (``achiasmy''), while the other sex recombines (Burt
et al., 1991; Haldane, 1922; Huxley, 1928). Alternatively, in most
species with recombination, crossovers occur in both sexes but at
different rates (``heterochiasmy''). In these species, females tend to
recombine more than males (Bell, 1982; Brandvain and Coop, 2012; Burt et
al., 1991; Lenormand and Dutheil, 2005; Lorch, 2005). In plants,
heterochiasmy is correlated with the opportunity for haploid selection
(Lenormand and Dutheil, 2005). Despite the establishment of these
interspecific trends, an understanding of how sex shapes the evolution
of recombination cannot be achieved with available data.

Comprehensive comparisons of variation in female and male recombination
rates within species have come from outbred populations of humans (Gruhn
et al., 2013; Halldorsson et al., 2019; Kong et al., 2004, 2014, 2008),
dog (Campbell et al., 2016), cattle (Ma et al., 2015; Shen et al.,
2018), and Soay sheep (Johnston et al., 2016), in which the role of sex
is confounded with the contributions of genetic variation. Although it
is known that the level and direction of heterochiasmy can differ among
species (Brandvain and Coop, 2012; Lenormand and Dutheil, 2005), the
correlation between female and male recombination rates among closely
related species remains poorly documented. Direct contrasts between the
two sexes across a common, diverse set of genomic backgrounds that
represent recent timescales would reveal whether the genome-wide
recombination rate evolves differently in males and females.

Examining variation in the total number of crossovers in a sex-specific
manner could also illuminate evolutionary connections between
recombination rate and crossover positioning. Analyses of meiotic
chromosome morphology in \emph{Arabidopsis thaliana},
\emph{Caenorhabditis elegans}, and \emph{Mus musculus} suggest that the
sex with more recombination usually has longer chromosome axes (Cahoon
and Libuda, 2019). A survey of 51 species found conserved sex
differences in the recombination landscape, including telomere-biased
placement of crossovers in males but not in females (Sardell and
Kirkpatrick, 2020). The degree to which a crossover reduces the
probability of another crossover nearby (crossover interference) also
differs between females and males (Otto and Payseur, 2019).

The house mouse, \emph{Mus musculus}, is a compelling system for
understanding how sex affects the evolution of recombination. Multiple
subspecies share a most recent common ancestor approximately 0.5 million
years ago (Geraldes et al., 2011), providing the opportunity to examine
natural variation on recent evolutionary timescales. Wild \emph{Mus
musculus} belong to the same species as classical inbred strains of
mice, where the molecular and cellular pathways that lead to crossovers
have been studied extensively (Baudat et al., 2013; Bolcun-Filas and
Schimenti, 2012; Handel and Schimenti, 2010). Single-cell
immunofluorescent approaches make it possible to estimate genome-wide
recombination rates in individual males and females (Koehler et al.,
2002; Peters et al., 1997). A collection of wild-derived inbred strains
founded from a variety of geographic locations is available, enabling
genetic variation in recombination to be profiled across the species
range. Most importantly, by measuring recombination rates in females and
males from the same set of wild-derived inbred strains, the evolutionary
dynamics of recombination can be directly compared in the two sexes.

In this paper, we report genome-wide recombination rates from both sexes
in a diverse panel of wild-derived inbred strains of house mice and
their close relatives. We demonstrate that recombination rate evolves
differently in females and males, even over short timescales.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth,height=\textheight]{Figure1.jpg}
\caption{Caption for the picture.}
\end{figure}

@@

We used a panel of wild-derived inbred strains of house mice (\emph{Mus
musculus}) and related murid species to profile natural genetic
variation in recombination (Table 4). Mice from the same inbred strain
served as serve as a biological replicates. Our survey included 5
strains from \emph{M. m. musculus}, 4 strains from \emph{M. m.
domesticus}, 2 strains from \emph{M. m. molossinus}, 2 strains from
\emph{M. m. castaneus}, and 1 strain each from \emph{M. spicilegus},
\emph{M. spretus} and \emph{M. caroli}. We subsequently denote strains
by their abbreviated subspecies and name
(\emph{e.g.~domesticus\textsuperscript{WSB}}). Mice were housed at
dedicated, temperature-controlled facilities in the UW-Madison School of
Medicine and Public Health, with the exception of mice from Gough
Island, which were housed in a temperature-controlled facility in the
UW-Madison School of Veterinary Medicine. Mice were sampled from a
partially inbred strain of Gough Island mice, approximately 6
generations of brother-sister matting. All mice were provided with
\emph{ad libitum} food and water. Procedures followed protocols approved
by IACUC.

@@

The same dry-down spread technique was applied to both spermatocytes and
oocytes, following Peters et al. (1997), with adjustment for volumes.
Spermatocyte spreads were collected and prepared as described in
Peterson et al. (2019). The majority of mice used for MLH1 counts were
between 5 and 12 weeks of age. Juvenile males between 12 and 15 days of
age were used for DMC1 counts. Both ovaries were collected from embryos
(16-21 embryonic days) or neonates (0-48 hours after birth). Whole
testes were incubated in 3ml of hypotonic solution for 45 minutes.
Decapsulated ovaries were incubated in 300ul of hypotonic solution for
45 minutes. Fifteen microliters of cell slurry (masticated gonads) were
transferred to 80ul of 2\% PFA solution. Cells were fixed in this
solution and dried in a humid chamber at room temperature overnight. The
following morning, slides were treated with a Photoflow wash (Kodak,
diluted 1:200). Slides were stored at -20*C if not stained immediately.
To visualize the structure of meiotic chromosomes, we used antibody
markers for the centromere (CREST) and lateral element of the
synaptonemal complex (SC) (SYCP3). Crossovers (COs) were visualized as
MLH1 foci. Double strand breaks (DSBs) were visualized as DMC1 foci. The
staining protocol followed (Anderson et al., 1999) and (Koehler et al.,
2002). Antibody staining and slide blocking were performed in 1X
antibody dilution buffer (ADB) (normal donkey serum (Jackson
ImmunoResearch), 1X PBS, bovine serum albumin (Sigma), and Triton X-100
(Sigma)). Following a 30-minute blocking wash in ABD, each slide was
incubated with 60ul of a primary antibody master mix for 48 hours at
37*C. The master mix recipe contained polyclonal anti-rabbit anti-MLH1
(Calbiochem; diluted 1:50) or anti-rabbit anti-DMC1 (mix of DMC1),
anti-goat polyclonal anti-SYCP3, (Abcam; diluted 1:50), and anti-human
polyclonal antibody to CREST (Antibodies, Inc; diluted 1:200) suspended
in ADB. Slides were washed twice in 50ml ADB before the first round of
secondary antibody incubation for 12 hours at 37*C. Alexa Fluor 488
donkey anti-rabbit IgG (Invitrgoen, location; diluted to 1:100) and
Coumarin AMCA donkey anti-human IgG (Jackson ImmunoResearch; diluted to
1:200) were suspended in ADB. The last incubation of Alexa Fluor 568
donkey anti-goat (Invitrogen; diluted 1:100) was incubated at 1:100 for
2 hours at 37* C. Slides were fixed with Prolong Gold Antifade
(Invitrogen) for 24 hours after a final wash in 1x PBS. Three slides of
cell spreads per mouse were prepared to serve as technical replicates
for the staining protocol. Comparisons of multiple stained slides from
the same mouse showed no difference in mean MLH1 cell counts and mean
cell quality. Sampled numbers of mice and cells per mouse were maximized
to the extent possible given constraints on breeding and time.

@@

Images were captured using a Zeiss Axioplan 2 microscope with AxioLab
camera and AxioVision software (Zeiss, Cambridge, UK). The number of
cells imaged per individual mouse is based on previous studies (Dumont
and Payseur, 2011; Murdoch et al., 2010; Wang and Payseur, 2017).
Preprocessing, including cropping, noise reduction, and histogram
adjustments, was performed using Photoshop (v13.0). Image file names
were anonymized before manual scoring of MLH1 foci or DMC1 foci using
Photoshop.

\hypertarget{analyses}{%
\subsubsection{Analyses}\label{analyses}}

To estimate the number of crossovers across the genome, we counted MLH1
foci within bivalents, synapsed homologous chromosomes. MLH1 foci were
counted in pachytene cells with intact and complete karyotypes (19
acrocentric bivalents and XY for spermatocytes; 20 acrocentric bivalents
for oocytes) and distinct MLH1 foci. A quality score ranging from 1
(best) to 5 (worst) was assigned to each cell based on visual appearance
of staining and spread of bivalents. Cells with a score of 5 were
excluded from the final analysis. Distributions of MLH1 count per cell
were visually inspected for normality (Supplemental Figure 1). When
outliers for MLH1 count were found during preliminary analysis, the
original images were inspected and the counts confirmed.

MLH1 foci located on the XY in spermatocytes were excluded from counts.
In addition to MLH1 counts, we measured several traits to further
characterize the recombination landscape. To estimate the number of
double-strand breaks, a minority of which lead to crossovers, mean DMC1
foci per cell was quantified for a single male from each of a subset of
strains (\emph{molossinus\textsuperscript{MSM}},
\emph{musculus\textsuperscript{PWD}},
\emph{domesticus\textsuperscript{WSB}}, and
\emph{domesticus\textsuperscript{G}} ). SC morphology and CREST foci
number were used to stage spermatocytes as early zygotene or late
zygotene.

To measure bivalent SC length, two image analysis algorithms were used.
The first algorithm estimates the total (summed) SC length across
bivalents for individual cells (Wang et al., 2019). The second algorithm
estimates the SC length of individual bivalents (Peterson et al., 2019).
Both algorithms apply a `skeletonizing' transformation to synapsed
chromosomes that produces a single, pixel-wide `trace' of the bivalent
shape. Total SC length per cell was quantified from pachytene cell
images (Wang et al., 2019).

To reduce algorithmic errors in SC isolation, outliers were visually
identified at the mouse level and removed from the data set. Mouse
averages were calculated from cell-wide total SC lengths in 3,195 out of
3,871 cells with MLH1 counts. SC length of individual bivalents was
quantified in pachytene cell images (Peterson et al., 2019). The DNA
CrossOver algorithm (Peterson et al., 2019) isolates single,
straightened bivalent shapes, returning SC length, location of MLH1
foci, and location of CREST (centromere) foci. The algorithm
substantially speeds the accurate measurement of bivalents, but it
sometimes interprets overlapping bivalents as single bivalents. In our
data set, average proportions of bivalents per cell isolated by the
algorithm ranged from 0.48 (\emph{molossinus\textsuperscript{MSM}} male)
to 0.72 (\emph{musculus\textsuperscript{KAZ}} female). From the total
set of pachytene cell images, 10,213 bivalent objects were isolated by
the algorithm. Following a manual curation, 9,569 single-bivalent
observations remained. The accuracy of the algorithm is high compared to
hand measures after this curation step (Peterson et al., 2019). The
curated single bivalent data supplemented our cell-wide MLH1 count data
with MLH1 foci counts for single bivalents. Proportions of bivalents
with the same number of MLH1 foci were compared across strains using a
chi-square test.
